Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-08-07 Manuscript/manuscript.tex
Ca2+ electrometer results from a -close-to-final- model of Lipid14-based POPC (q_sc=0.8, sigma_sc=0.875, all non-tail atoms), extra scripts for GMX deffnm simulations, Manuscript: title of the future paper.
Updates to plots: OP-change plot has corrected concentrations on x-axis; added experimental values for tails to OrdPar plot with actual values;  updated density plots; Added simulation data for model f_q=0.8, f_siq=0.89; debugged scripts for order-parameter calculation added simple pythonic script for the calculation of corrected ionic concentration.

\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}

@@

\title{NMRlipids VI: New PC lipid model with correct cation binding affinity} %Title of paper

@@

\begin{abstract}
% insert abstract here
  Cellular membranes are complex systems that are hard to study both experimentally and theoretically. Classical molecular dynamics simulations (MD)
give detailed information about membrane structure, dynamics and its interactions with other biomolecules. However,  there is still a room for
improvements in the current force fields, especially in the hydrophilic region of membrane.  According to the results reported in NMRlipids, Open Collaboration project
running at nmrlipids.blogspot.fi, biologically relevant Na+ and Ca2+ generally overbind to zwitterionic phophatidylcholine (PC) membranes in current MD models [Ollila 2016].
Here we suggest that the PC lipid membrane-ion interactions can be corrected by implicitly inducing electronic polarizability in the lipid models by
using a simple electrostatic continuum correction (ECC) for partial charges [Leontyev 2010], which is successfully applied already to monovalent and divalent ions in bulk water [Kohagen, 2015].
Our results show that applying ECC to existing force fields significantly improved PC lipid--cation interactions when compared against NMR data
by using the electrometer concept [Ollila 2016,Seelig87]. However, also other force field parameters have to be made consistent
for ECC correction to preserve the area per molecule and other relevant bilayer properties. This work is carried as Open Collaboration
in NMRlipids VI project (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

%\label{}
\section{Introduction}

Cations were shown to generally overbind in PC lipid bilayers in
NMRlipids II project. Here we propose that the cation overbinding can
be corrected by implicitly inducing electronic polarizability in
lipid headgroups by scaling the partial charges.

%\section{Methods}

%\subsection{Molecular dynamics simulations}

\section{Effect of scaling of headgroup and glycerol backbone partial charges on lipid bilayer properties and cation binding}

Headgroup and glycerol backbone partial charges of Slipids and Lipid14 models were scaled
with various scaling factors and cation binding was monitored by using the electrometer
concept. Headgroup order parameter changes as a function of ion concentration are
shown in Fig. \ref{OrderParameterIONSchangesSCALED}. The order parameter decrease
seemst to be proportional to the headgroup charges: the smaller the charges
(larger scaling factor), the less change is observed in order parameters.
According to the electrometer concept, this means that the headgroup scaling
decreases cation binding affinity. However, from this data only we cannot
exclude the possibility that order parameters are changing less because
headgroup becomes less sensitive for cation binding.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/OrderParameterIONSchangesSCALED.eps}
  \caption{\label{OrderParameterIONSchangesSCALED}
    Headgroup order parameter changes as a function of cations from models
    with modified headgroup partial charges.}
\end{figure}

The best models from Fig.  \ref{OrderParameterIONSchangesSCALED} are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}. Based on these results we choose
two models for more careful studies: Slipids and Lipid 14 models with headroup
and glycerol backbone partial charges scaled with 0.8 and 0.85, respectively.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/OrderParameterCHANGESnewMODELS.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of cations from models
    with modified headgroup partial charges.}
\end{figure}

While scaling seems to improve ion binding, it changes the lipid bilayer properties
without ions. The glycerol backbone and headgroup order parameters are plotted in
Fig \ref{HGopsNEWmodels.eps} for reference models and EEC models. The difference,
i.e. the effect of scaling on these parameters, is shown in Fig. \ref{HGopsNEWmodelsCHANGE}.
The area per lipid from different models are shown in Table \ref{apls}
The conclusion is that the charge scaling has only a marginal effect on
headgroup order parameters, while area per lipid is significantly decreased.
Area per lipid decrease is more pronounced in Lipid14 than in Slipids.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/HGopsNEWmodels.eps}
  \caption{\label{HGopsNEWmodels}
    Headgroup and glycerol backbone order parameters from standard and EEC models.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/HGopsNEWmodelsCHANGE.eps}
  \caption{\label{HGopsNEWmodelsCHANGE}
    Changes in headgroup and glycerol backbone order parameters due to EEC.}
\end{figure}

\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{apls} }
  \begin{tabular}{c c}
    model          & A (Å$^2$)    \\
    Lipid14 (literature)  & 65.6$\pm$ 0.5   \\
    Lipid14eec0.85 & 55.5      \\
    Slipids (literature T=303K)       & 64.6 $\pm$ 0.4   \\
    SlipidsEEC0.8  & 57.8  \\
  \end{tabular}
\end{table}

The Calcium densities from new models and standard models are compared in Fig \ref{CAdensities}.
It seems to me that Ca binding in Lipid14 model is not reduced by scaling the
partial charges even thought the order parameter response is better.
On the other hand, in Slipids the binding seems to be reduced.
Possible interpretation is that Lipid14 result gets better because headgroup
gets less sensitive for bound charge, while in Slipid the binding is reduced.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/CAdensities.eps}
  \caption{\label{CAdensities}
    Calcium densities from different simulations with new and standard models.
  }
\end{figure}

\section{Conclusions}

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

>>>>

\documentclass[aip,jcp,twocolumn]{revtex4}
%\documentclass[aip,jcp]{revtex4}
%\documentclass{article}

@@

\title{Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization} %Title of paper

@@

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and head group order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids VI (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} \todo{We need to select references for these}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding model and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~\ce{Ca^2+} induced charge inversion of bacterial membranes~\cite{Luan2016}).

%The binding details, like binding sites and stoichiometry are not yet fully
%resolved but interpretation of NMR and scattering experiments suggest that one
%Ca2+ interacts mainly with the choline groups \cite{hauser76,hauser78,herbette84} of two
%phospholipid molecules \cite{altenbach84}.

In this work we show that the accuracy of ion binding, especially calcium,
in zwitterionic PC lipid bilayer can be significantly improved 
for classical MD simulation models of zwitterionic lipids
by including electronic polarizability
in the polar region of lipid molecules
using the electronic continuum correction (ECC)~\cite{leontyev11}. 
This approach has been previously shown to improve performace of
ion models against neutron scattering data in concentrated solutions \cite{Jungwirth2015,kohagen14,kohagen16}
and ab-initio simulations~\cite{Jungwirth2017}. 
The realistic structure of ions in a bulk solvent was not, however, sufficient to correct binding in lipid
bilayers~\cite{catte16}. 
To fix it, we propose
a classical fixed-charge MD simulation model of 
1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) lipid 
that accounts for electronic polarizability by using ECC. 
We will validate such a model by comparing its lipid bilayer structue
to x-ray scattering data and its head group structure to NMR experiments. 
We will then quantify the accuracy of head group order parameter response to bound charge
and compare it to the experimental NMR data of sodium and calcium binding in PC lipid bilayers. 
In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.
The proposed POPC model, ECC-lipids, and the approach to obtain it
is, hence, highly useful for future MD simulations with physiological salt conditions, 
and for effects where membrane electrostatics play a major role.

\section{Methods}

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability in the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations. Some pioneering simulation studies scaled
the partial charges of atoms by one half to effectively include 
polarizability \cite{jonsson86,egberts94}. Also, approaches that
explicitly include polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
In contrast to those studies, however, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

The main goal of this work is to apply ECC correction to accurately 
describe the lipid head group response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16, Pluharova2014},
it is not clear if the same factor should be used for lipid head groups. 
Unlike the total charge of cations and anions, atomic partial charges used in
classical MD simulations are not experimental observables and 
can in principle adopt arbitrary values. 
Here we estimate the correct scaling factor for Lipid14 model by using
so called ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013},
which are the average of partial charges given by 
restrained electrostatic potential fitting (RESP~\cite{RESP_paper}) 
in vacuum and in solvent.
Using the partial atomic charges obtained from RESP fitting in vacuum and implicit solvent reported in \cite{maciejewski14} and
considering that the partial charges in original Lipid14 parametrization were derived using RESP~\cite{RESP_paper} in vacuum,
we conclude that the charges in Lipid14 model shall be scaled roughly by the factor $f_q=0.8$. 
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

Here we apply the ECC correction with the proposed scaling factor only to the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized in most lipids models including Lipid14 and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16}.
In contrast, all available lipid models require
some improvements in glycerol backbone and head group regions \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications. 
The scaling of partial charges in the polar region was found to decrease
area per molecule to 60~\AA$^2$, which is significantly smaller than the
experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment}) and the original Lipid14 values $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}..  
The decrease of area was found to arise from a lower hydration of the lipid head group region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. 
The solvation free energy and hydration is sensitive to Van der Waals interaction parameters,
hence, they can be increased by reducing the radius of the selected atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential. 
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges were scaled. This has lead to an increase of the area per molecule back
close to the experimental value (see Table~\ref{tab:apls}). 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that the sensitivity of the order parameters response
to the bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. Hence, we first quantify the response of
head group order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of the bound charge is known
and can be plotted against order parameter change.
Secondly we evaluate the response of the head group order parameters to varying salt concentrations in solution
and estimate the amount of membrane bound charge.

%\todo{ongoing,Actual concentration of cations in simulation has yet to be estimated. 
%If it varies too much from the nominal concentration, I may need to tweak the scaling 
%factors, $f_q$ or only $f_\sigma$, to accomodate it. However, it is very unlikely, response 
%to the surfactant \ce{DHMDMAB} is OK. Big patches with loads of solvent are running at the 
%moment to guide me on the possible finite-size errors and this conc-error.
The used experimental data report order parameters as a function of
equilibrium cation concentration in the bulk solvent \cite{akutsu81,altenbach84}.
Such a condition is reached in simulations by adjusting
the simulation box size to dimensions large enough
that ion conceration reaches a clear plateau in the bulk solvent.
The concentrations in the units of $\mathrm{mol/l}$ were then 
determined as
\begin{equation}
  C_{eq}=\frac{C_{plateau}}{0.602},
\end{equation}
where plateau concentration is the number density in the units of $\mathrm{nm}^{-3}$.
\todo{SAMULI: Once we have to final results, we can probably say that
  the repeat distance is not far from the experimentally measured distance \cite{pabst07,uhrikova08}
  }

\subsection{Comparison of lipid bilayer structure to experimental data}

Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}. Form factor was calculated from simulations
as 
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

\subsection{Simulation details}

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

\section{Results and Discussion}

\subsection{POPC membrane and its structure}

\begin{figure}[tbp]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure} \\
  \todo{We should run a simulation for the system withtout ions with the
    same (or close) temperature as the experimental data. The NMR data is at 300K
    so that would be good. I think that this would be also close enough for the scattering data at 303K.
  }
\end{figure}

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC without ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids &        &  \\
    %Lipid14ecc0.80+sigma0.875 &        &  313    \\
    ($4.6\cdot 5.1 \, \mathrm{nm}^2$), 72 lipids patch, OPC3           & 63.2   &   313      \\
    \hline
    (6.4 nm)$^2$, 128 lipids patch, OPC3           & 64.2   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, SPC/E          & 65.1   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, OPC            & 64.4   &  313       \\
    (6.4 nm)$^2$, 128 lipids patch, TIP4p/2005     & 66.8   &  313       \\
    %($9.2\cdot 10.2 \, \mathrm{nm}^2$), 288 lipids patch           & 65.5   &  313       \\ %% not done for this model with f_sigma=0.89
    %oMM small patch           & 63.65  &         \\
    %oMM 4xbig patch           & 63.7   &         \\
    \hline
    %experiment   & 62.7  &  293    \\
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    %experiment  & 68.1  &  333    \\
    %experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

In order to validate the newly developed model, ECC-lipids, 
we compared our simulation results without ions to
NMR order parameters measurements and x-ray scattering form factors 
(Fig.~\ref{simVSexpNOions} and Table~\ref{tab:apls}). 
The tail order parameters being already highly optimized in the original Lipid14 model~\cite{dickson14}
are found to match the experimental values nicely.
The head group and glycerol backbone order parameters accuracy is 
comparable to the state of art lipid models available in literature \cite{ollila16},
The agreement between the x-ray scattering form factors and areas per molecule from simulations and experiments
confirm that the membrane structural properties are well captured. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

\subsection{POPC membrane and electrometer concept with cationic surfactant}

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2_surf.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Headgroup order parameter changes as a function of NaCl, CaCl$_2$ concentration and
    cationic surfactant (dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) mole fraction
    from simulations and experiments (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, surfactant \cite{scherer89}).
  }
\end{figure*}

Headgroup order parameter response to the bound charge was evaluated against
experimental data measured with the cationic surfactant
(dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) \cite{scherer89}.
The exact amount of bound charge in the membrane is known in these systems, because
practically all cationic surfactant molecules are embedded in the lipid bilayer due to
their amphiphilic nature. Thus, the changes of head group order parameter as a function of
mole fraction of cationic surfactants (Fig. \ref{OrderParameterCHANGESnewMODELS}) gives
also the order parameter changes as a function of the amount of bound cations.
\todo{Split the OP-changes figure in two: cationic surfactant -and- binding of cations.
      The separate figure with the cationic surfactant shall be additionally accompanied
      with a plot of Head group tilt (PN-vector?) change/response. }

The head group order parameter response to the bound cation concentration
is in experiments approximately linear up to $\sim$0.3 mole fraction of the cation \cite{scherer89}.
The linearity is also observed in simulations with original Lipid14 and with newly derived ECC-lipids models.
Quantitative comparison, however, reveals that the response is overestimated in 
Lipid14 for both segments, while the ECC-lipids model gives a
good agreement for the change of $\alpha$ segment order parameter, and slightly
underestimates the $\beta$ segment order parameter change.
Such a finding is not surprising, as the $\Delta S^\alpha/\Delta S^\beta$ ratio 
of the response of Lipid14 model is found to be slightly larger than in 
experiments (VALexp vs VALsim
\todo{Add values of $S^\alpha/S^\beta$ response for sim and experiment
  SAMULI: Maybe these could be put in the same figure with the order parameter
  response and P-N vector angle change with surfactants.
}
), and this property is also kept by the newly derived Lipid14-based model, ECC-lipids. 
The present comparison of head group order parameter changes 
suggests that the observed overestimated response of all models to 
increasing CaCl$_2$ concentration in \cite{catte16} 
can be explained at least in part by a high sensitivity of the head group order parameter response
to the bound charge.

\subsection{POPC membrane and electrometer concept with cation binding}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities.eps}
  \caption{\label{fig:cacl-dens}
    Density profiles of \ce{Ca^{2+}} and \ce{Cl^-} for Lipid14 model with Dang ions
    and with ECC-ions and ECC-lipids model with ECC-ions. }
  \todo{Change the figure so that it contains a membrane background} \\
  \todo{What are the concentrations (calculated as in NMRlipids II) in this figure?} \\
  \todo{These should be also compared to the results with Åqvist ions, which are shown
    in NMRlipids II. NMRlipids II have results with 350mM and 1000mM. If we have the distributions with
    the same concentrations Åqvist results could be added also in this figure. If not, the
  comparison has to be made some other way.}
\end{figure}

After achieving accurate head group order parameter response with cationic surfactant, 
we advanced to the response to solutions with varying concentrations of salts, namely \ce{NaCl} and \ce{CaCl_2}. 
Headgroup order parameter response to increasing CaCl$_2$ concentration
from experiments, original Lipid14 model and ECC-lipids model are shown
in Fig. \ref{OrderParameterCHANGESnewMODELS}. The order parameter response
is significantly overestimated in original Lipid14 model, while results from
ECC-lipids model are in a good agreement with the experiments.
This is a significant improvement over current models,
which in general overstimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicate
that the binding details of \ce{Ca^2+} are realistic in the ECC-lipids model
and, hence, it can be used to study lipid-ion interaction in detail.

Ion density profiles from simulations with different models are compared in Fig.~\ref{fig:cacl-dens}.
The density profiles from simulations with original Lipid14 \cite{dickson14} and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of a ECC-ion model~\cite{kohagen14,kohagen16} 
or the standard Amber ion model by \AA{quist} 
along with Lipid14 does not significantly change it 
(Headgroup order parameter responses for these models in SI).
\todo{Add OP-response of Lipid14+ECC-ions into the OP-changes plot in SI}
The new ECC-lipids model with ECC-ions exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
Together with the accurate head group order parameter response (Fig.~\ref{OrderParameterCHANGESnewMODELS}),
we can conclude that the present model, ECC-lipids, exhibits accurate ion binding affinity to zwitterionic phospholipid membranes
suggesting that the models that overestimate head group order parameter response~\cite{catte16} also overestimate the amount of bound cations.

\begin{figure}[]
  \centering
  \includegraphics[height=9.0cm,angle=-90]{../Fig/bound-CAs_conc-eccl17.eps}
  \caption{\label{fig:cacl-bind}
    Binding isotherm assuming stoichiometry of 2~POPC:1~\ce{Ca^{2+}} as used in \cite{altenbach84} fits our simulation data well .}
\end{figure}

Good agreement of ECC-lipids model with experiments encourages us to analyse the
binding details from MD simulations. Direct analysis of contacts between ions and
lipids from simulations suggest that the most abundant POPC:\ce{Ca^2+} complex 
have stoichiometry of 2~POPC:1~\ce{Ca^{2+}}.
As shown in Fig.~\ref{fig:cacl-bind} this is in agreement
with the ternary complex model suggested based on head group order parameter
experiments~\cite{altenbach84}.
In addition to the ternary complexes, there also is a non-negligable probability
of one \ce{Ca^{2+}} cross-bridging three POPC molecules.
Technical details of the analysis are in the SI.
